Higher Education Synthesized

### 2025-03-18 artifacts/report/msds_data_mining_project_report.tex
update conclusion section of report

\section{Discussion}

    Do we generate one larger dataset and then push that into the warehouse? Do we generate a bunch of smaller datasets knowing that the cross-section between them earlier in the pipeline may get lost?
    Appropriate and responsible ways to use generative modeling?
    Could an individual institution take this work and use it in tandem with their internal student-level data to get a personalized picture?

    In discussion section, emphasize that we did spend a lot more time on generative modeling and cut scope after to leave space for generative modeling exploration
    In discussion, include what lessons we learned in this project

    Should we train generative models on smaller or combined datasets? Do we lose anything by training them on smaller datasets later in the pipeline? Depends on use case. If we want the warehouse to have synthetic data, then we should model earlier. If we are using synthetic data only later for analysis, then maybe it doesn't matter as much.

    My goal for this project is to stick within a 3-4 week limit, if possible. I tend to drag out projects, taking the opportunity to add increased complexity and explore new items I am curious about. But, while spending forever on projects is fun and educational, I do risk getting too comfortable with taking my time. I could benefit from descoping and focusing on getting smaller chunks working earlier and then deciding if it is worth expanding out from there.

    The first area to keep an eye on scope is with the data used in the project. We start by selecting three IPEDS domains among many. We then systematically trim the datasets down in stages, as described earlier. By the time we reach modeling, we have a dataset with 14 features. Two of those are numeric targets we can use for regression. We could also put some boundary cutoffs for those fields to aid with classification. We have a couple of more likely features in terms of explanatory power, and we have a few extras to test out in terms of importance. Keeping the dataset smaller helps keep the modeling stages more focused.

    Next is the use of the database. This is a personal thing for me, but I am more familiar working in databases over the years, so doing work in the database simplifies the amount of time needed for me to set up a production-esque system.

    I also selected data that is complex but familiar. I have worked in higher education for years, so, while IPEDS data is itself complicated, it helps that I am used to using higher education data, again helping to keep the scope manageable.

    The result of trimming the scope is that I leave more time to play with generative modeling. While the final generative modeling notebook is pretty streamlined, it did not start that way. I had the opportunity to explore a number of models spanning different algorithms and a range of hyperparameters.

    My initial main challenge that I was looking out for was much higher complexity in the IPEDS data. To be fair, the IPEDS data took a good chunk of the time in this project to sort through, and, as stated earlier, I purposely narrowed down the data to try and address this challenge.

@@

    In terms of ongoing challenges, I am currently working through evaluations of the synthetic data. I have refined the process quite a bit, but there are still some remaining areas to smooth out. For example, I already had to drop some features due to their negative performance on generative model training. I would like to explore ways to tinker with the generative modeling to encompass a broader range of data. This could look like changing hyperparameters, but it is more likely that it means curating different modeling datasets to answer specific questions.

    I already leveraged one of my backup plans in that I needed to descope the range of IPEDS data that I originally looked to include. Three separate times.

    If I run into any serious issues with generative modeling, it is an option to test out different libraries. I have a few in mind based on other research papers I have worked through in the past, but I am not realistically expecting to switch libraries this far into the project.

    A final backup plan would be what to do if the steps after generative modeling end up a bit too plain. For example, while the synthetic data we create now may look good on the surface, it may not be as useful as we would like it to be in terms of regression, classification, or other approach. The simpler plan here would be to expand the evaluation steps for the synthetic data and generative modeling since this already includes some extra modeling. It could be an option to return to the IPEDS files and select a wider range of features, but I would need to see some serious trouble before wanting to shift back that far in the pipeline to start redoing work.

    I do want to call out that, since this report comes most of the way through the planning, design, and implementation of the project, I am hoping that there are not too many surprises left, if any. I have also already taken steps to mitigate my biggest worry of scope creep, so I am not as worried about that creeping back in at this point. It will require not letting final modeling take up too much space, but, fortunately, the project feels pretty robust as it is now, so there may be an opportunity to find a way to wrap things up soon.

>>>>

\section{Discussion}
    My goal for this project was to stick within a 3-4 week limit, and I was able to stick within this timeframe. I tend to drag out projects, taking the opportunity to add increased complexity and explore new items I am curious about. But, while spending forever on projects is fun and educational, I do risk getting too comfortable with taking my time. I could benefit from descoping and focusing on getting smaller chunks working earlier and then deciding if it is worth expanding out from there.

    The first area in which to keep an eye on scope is with the data used in the project. We start by selecting three IPEDS domains out of many. We then systematically trim the datasets down in stages, as described earlier. By the time we reach modeling, we have a dataset with 14 features. Two of those are numeric targets we can use for regression. We could also put some boundary cutoffs for those fields to test out classification. We have a couple of more likely features in terms of explanatory power, and we have a few extras to test out in terms of importance. Keeping the dataset smaller helps keep the modeling stages more focused.

    I also selected data that is complex but familiar. I have worked in higher education for years, so, while IPEDS data is itself complicated, it helps that I am used to using higher education data, again helping to keep the scope manageable. There are a lot of interesting higher education questions that got dropped while trimming the data down, but it will be feasible to add those back in in future iterations based on the design of the system.

    So, I ended up leveraging one of my backup plans in that I needed to descope the range of IPEDS included in the project. Three separate times.

    Next is the use of the database. This is a personal thing for me, but I am more familiar working in databases over the years, so doing work in the database simplifies the amount of time needed for me to set up a production-esque system. We also run into databases everything in real-world systems, so a project that mirrors a production pipeline, even if on a very small scale, must have database tech included in it, in my experience.

    Finally, a big area of scope reduction is at the end of the pipeline. I originally intended to implement both generative modeling and analytical modeling. Generative modeling is key to the focus of this project, but I was unclear if the project would feel incomplete if it did not have extra layers of analytical modeling as well. It was not until this final update for the project and deliverables that I decided to stick with the generative modeling and leave the analytical modeling until a future expansion of the project. Since I decided to explore generative modeling early in the planning stages, that took priority over analytical modeling, and the generative modeling itself took a significant chunk of time to familiarize myself with, experiment with, and implement. To add extra analytical modeling at this stage feels like it would balloon the scope and pull the focus of the project elsewhere.

    Speaking of generative modeling, even after all the tinkering I did for this project, I still have some lingering questions. The first is where the generative modeling should happen. The approach in this project was to pull the analysis dataset out of the warehouse and synthesize more data mimicking that dataset. If we zoom out and think about a pipeline and warehouse that needs to be more general purpose, there is a case to synthesize data earlier, possibly even keeping real and synthetic fact and dimension tables in the warehouse. If the goal of the synthetic data is to distribrute a final dataset, then our approach here works well. If the goal is to incorporate synthetic data as a core piece of the warehouse, then we may want to synthesize data earlier.

    I am still on the fence about the right answer here. I think the location of generative modeling in this pass of the project is correct, but the big downside is that we are generating new data based on a very small and bespoke subset of the overall IPEDS data that we start with in this project, and that is already a big reduction in terms of the overall IPEDS data that is available across more domains. With my current understanding of generative modeling, we may want the modeling to be able to look at a wider range of IPEDS data when learning patterns to synthesize since we may drop some important relationships between features before we hit generative modeling in this project. We could generate one much larger dataset and then push that into the warehouse, possibly adding generative modeling as a step between the EDA after the data lake and before transforming data into the dimensional model. The trade-off though is that it feels strange to add generative modeling in too early, like we are building a warehouse and pipeline primarily to serve synthetic data, not real data. This is one of my biggest questions that is left at the end of this iteration of the project.

    Another design question that has become more prominent as the project progressed has been how individual institutions may be able to combine their student-level data with the system here. As stated earlier, while synthesizing institutional-level data is valuable, it is likely that synthesizing student-level data is where the bigger value lies. One approach I can think of is to modify the pipeline to read in common student-level data, model it in the data lake and warehouse, and then prepare for it in terms of generative modeling, wherever we put that modeling step. I do not want to derail the report here with a lot of theorizing about how to incorporate student-level data, but it is worth noting that this is, along with where to plop the generative modeling, the biggest question I am left with at the end of the project.

    The initial main challenge that I was looking out for was much higher complexity in the IPEDS data. To be fair, the IPEDS data took a good chunk of the time in this project to sort through, and, as stated earlier, I purposely narrowed down the data to try and address this challenge.

@@

    In terms of ongoing challenges, I think the big one is keeping an eye on how to evaluate the synthetic data, especially if we start modifying the pipeline as described throughout this section. While I am comfortable with running evaluation metrics, interpreting them across different use cases will take a lot more experience. It would be good to put processes in place to track evaluation metrics over time and for different uses in order to improve the value and quality of the synthetic data. For example, we do not want synthetic data to 100\% match the real data, but we also do not want it to be so different that the conclusions you reach with the synthetic data are invalid.

    All-in-all, I am left with a number of questions and concerns at the end of this iteration of the project, a good sign that we accomplished some good work while also having room to enhance and expand the system going forward.
